---
title: "Model Card: Tidymodels machine learning model"
date: '`r Sys.Date()`'
output: 
  html_document
params:
    name: conformal_ensemble
    version: NULL
---

```{r, echo = FALSE, output=TRUE, eval=TRUE}
library(vetiver)
library(pins)
library(yardstick)
knitr::opts_chunk$set(echo = FALSE)

board <- board_s3(
  "data", 
  access_key = "user", 
  secret_access_key = "password", 
  region = "us-east-2", 
  endpoint = "http://127.0.0.1:9000"
)

v <- vetiver_pin_read(board, params$name, version = params$version)
v_meta <- pin_meta(board, params$name)

box::use(
  dials[svm_margin, trees],
  ggplot2[theme_set, theme_light],
  readr[read_csv],
  tune[control_bayes, extract_parameter_set_dials],
  here[here],
  ranger[ranger],
  kernlab,
  rsample[initial_split, vfold_cv],
  parsnip[rand_forest, svm_linear],
  dplyr[mutate],
  recipes[recipe, update_role, step_zv, step_nzv, step_normalize, step_impute_knn, update],
  yardstick[rmse, mae],
  hardhat[tune],
  workflows[workflow],
  workflowsets[workflow_set, option_add, option_add_parameters, workflow_map],
  stacks[stacks, add_candidates, blend_predictions, fit_members],
  vetiver[vetiver_model, vetiver_pin_write, vetiver_api],
  pins[board_temp, board_s3, pin_write, pin_read, pin_upload, pin_download],
  plumber[pr, pr_run],
  DALEXtra[explain_tidymodels]
)

theme_set(theme_light())

```

A [model card](https://doi.org/10.1145/3287560.3287596) provides brief, transparent, responsible reporting for a trained machine learning model.

## Model details

-   Developed by James Bristow
-   Name: tidymodel_ml_model
-   Version: 20251003
-   Framework: tidymodels
-   Target: mpg
-   Training Data: mtcars (32 rows, 11 columns)
-   If you have questions about this model, please contact [James Bristow](mailto:James.Bristow@plantandfood.co.nz){.email}

## Intended use

-   The primary intended uses of this model are for predicting miles per gallon for cars with similar weight and horsepower.
-   The primary intended users of this model are car enthusiasts.
-   Some use cases are out of scope for this model, such as for safety-critical or real-world fuel efficiency forecasting.

## Important aspects/factors

-   Predictors with most influence: weight (wt) and horsepower (hp).
-   Heavier cars result in lower mpg.
-   Higher horsepower result in lower mpg.
-   Data source: 1974 Motor Trend magazine road test (biased toward US cars of that era).
-   Scale: Very small dataset (n = 32), so model behavior is fragile.
-   Interpretability: Machine learning model may be challenging to directly interpret.

## Metrics

-   The metrics used to evaluate this model are root mean squared error and mean absolute error.
-   These metrics are computed via the [yardstick package.](https://yardstick.tidymodels.org/)
-   We chose these metrics because they are popular and interpretable evaluation metrics for predictive accuracy.

## Training data & evaluation data

-   The training dataset for this model was [*mtcard* from the 1974 Motor Trend magazine road test](https://cran.r-project.org/web/packages/explore/vignettes/explore-mtcars.html).
-   The training dataset for this model has the "prototype" or signature:

```{r, eval=TRUE}
dplyr::glimpse(v$prototype)
```

-   The evaluation dataset used in this model card is an 80-20 train-test data split of *mtcars*.
-   We chose this evaluation data because for demonstration purposes.

```{r, eval=TRUE}
df <- v$model$training
```

## Quantitative analyses {.tabset}

```{r, eval=TRUE}
library(parsnip)
library(workflows)
library(probably)
preds <- predict(v, df) |>
  dplyr::bind_cols(df)

preds  
```

### Overall model performance

```{r, eval=TRUE}
preds |>
    metrics(mpg, .pred)
```

### Visualize model performance

```{r, fig.height=3, eval=TRUE}
library(ggplot2)
p <- preds |>
    ggplot(aes(mpg, .pred)) +
    geom_abline(slope = 1, lty = 2, color = "darkblue", size = 1.2) +
    geom_point(alpha = 0.5, show.legend = FALSE) +
    geom_ribbon(aes(ymin = .pred_lower, ymax = .pred_upper), fill = "lightblue", alpha = 0.2)

plotly::ggplotly(p)
```

## Ethical considerations

-   Small, biased dataset (1974 US cars only).
-   Not representative of modern vehicles.

## Caveats & recommendations

-   This model is intended to be used with *mtcars* only.
-   This model doesn't generalize outside *mtcars*.
-   Interpret results as illustrative, not prescriptive.
-   We recommend fitting a similar model to a larger and newer dataset.
