---
title: "Machine Learning Workflow"
date: today
authors: "James Bristow, Jingjing Zhang, Hana Liang, Lindy Guo"
fig-pos: 'H'
toc: true
lof: true
lot: true
link-citations: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction

In this section, we will demonstrate how one may train a machine
learning model and tune its hyper-parameters using the *tidymodels*
framework, which is composed of several packages for implementing machine learning workflows
using tidyverse principles. *tidymodels* offers a unified interface to
performing data preprocessing, modelling using various machine learning
algorithms, model evaluation and visualisation, and cross-Validation &
tuning. [For more information, click this
link.](https://www.tidymodels.org/)

We first load all required dependencies.

```{r setup, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, output=TRUE, eval=TRUE}
box::use(../../R/proj_utils)
box::use(../../R/tidymodels_deployment)
box::use(
  dials[svm_margin, trees, penalty, mixture],
  readr[read_csv, read_file],
  tune[control_grid, extract_parameter_set_dials, .get_tune_metric_names, select_best],
  here[here],
  dplyr[mutate, bind_rows, arrange, select, pull, slice_head],
  purrr[pluck],
  ranger[ranger],
  kernlab,
  rsample[initial_split, vfold_cv, training, testing],
  parsnip[rand_forest, svm_linear, boost_tree, set_engine, set_mode, linear_reg],
  recipes[recipe, update_role, step_zv, step_nzv, step_normalize, step_impute_knn, step_rm, update, step_pca, prep, bake, all_numeric, all_outcomes, all_numeric_predictors],
  yardstick[rmse, mae, metric_set, metrics],
  hardhat[tune],
  ggplot2[ggplot, geom_point, labs, theme_minimal, aes, geom_ribbon, geom_col, coord_flip, facet_wrap],
  workflows[workflow, add_recipe, add_model],
  workflowsets[workflow_set, option_add, option_add_parameters, workflow_map, fit_best, workflow_autoplot = autoplot, rank_results],
  tidyclust[k_means, tidy_clust = tidy],
  stacks[stacks, add_candidates, blend_predictions, fit_members, stacks_autoplot = autoplot],
  vetiver[vetiver_model, vetiver_pin_write, vetiver_api, vetiver_pin_read],
  pins[board_temp, board_local, board_s3, pin_write, pin_read, pin_upload, pin_download],
  paws.common,
  paws.storage,
  plotly[ggplotly],
  probably[int_conformal_split, cal_plot_regression],
  plumber[pr, pr_run],
  DALEX[feature_importance, model_profile, predict_profile, model_diagnostics],
  DALEXtra[explain_tidymodels],
  glmnet,
  zeallot[...],
  targets[tar_make]
)

source(here("R/probably_vetiver.R"))
```

# Data

We will next load an example dataset. [For this tutorial, we will use
the mtcars
dataset.](https://cran.r-project.org/web/packages/explore/vignettes/explore-mtcars.html)

The mtcars dataset is a classic dataset in R, extracted from the 1974 Motor Trend magazine. It contains data on fuel consumption and 10 aspects of automobile design and performance for 32 car models. 

- mpg: Miles per gallon, a measure of fuel efficiency.  
- cyl: Number of cylinders in the engine.  
- disp: Engine displacement in cubic inches.  
- hp: Gross horsepower of the car.  
- drat: Rear axle ratio.  
- wt: Weight of the car (in 1000 lbs).  
- qsec: 1/4 mile time in seconds (acceleration measure).  
- vs: Engine shape (0 = V-shaped, 1 = straight).  
- am: Transmission type (0 = automatic, 1 = manual).  
- gear: Number of forward gears.  
- carb: Number of carburetors.

```{r load-tidymodels-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

data("mtcars")
mtcars <- mtcars |>
  mutate(car = rownames(mtcars))
head(mtcars)
```

We will next perform a train-test data split. The purpose of data splitting is to fairly evaluate how well a predictive model will perform on new, unseen data. We don’t want to just measure performance on the data used to train the model because the model might just memorise it (overfitting).

We will use an 80-20 train-test data split. We will train our model on 80% of the data, and assess it against the other 20%.

```{r split-tidymodels-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
set.seed(200)
data_split <- mtcars |>
  initial_split(
    prop = 0.8
  )

data_train <- training(data_split)
data_test  <- testing(data_split)

head(data_train)
```

Next, we will create a data preprocessing recipe using the [recipes
package](https://recipes.tidymodels.org/), which is used to specify
sequences of feature preprocessing steps. This recipe will:

-   Drop zero and near zero-variance columns.
-   Normalise all numeric columns using Z-standardisation.
-   Impute missing data using k nearest neighbor imputation.

```{r process-tidymodels-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
training_recipe <- data_train |>
  recipe() |>
  update_role(everything(), new_role = "predictor") |>
  update_role(mpg, new_role = "outcome") |>
  step_zv() |>
  step_nzv() |>
  step_normalize(all_numeric(), -all_outcomes()) |>
  step_impute_knn()
```

# Exploratory Data Analysis: Dimensionality Reduction and Clustering

## Principal Component Analysis 

We will demonstrate how one may perform dimensionality reduction and clustering, primarily for visualisation purposes, on the mtcars dataset using tidymodels. We can take the data preprocessing training recipe specified above and extend from it to include a principal component analysis (PCA) preprocessing step. [Click this link for more information on PCA.](https://www.nature.com/articles/s43586-022-00184-w)

```{r tidymodels-pca-cluster, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Principal Component Analysis on mtcars
pca_prep <- training_recipe |>
  step_pca(all_numeric_predictors(), num_comp = 2) |>
  prep()

pca_results <- pca_prep |>
  bake(new_data = data_train) |>
  mutate(car = data_train$car)

pca_explained <- proj_utils$pca_var_explained(pca_prep, 2, 5) |>
  round(2) 
  
p <- pca_results |>
  ggplot(aes(PC1, PC2, label = car)) +
  geom_point(size = 1) +
  theme_minimal() +
  labs(x = paste0("PC1 (", pca_explained[1], " variance explained)"), y = paste0("PC2 (", pca_explained[2], " variance explained)"))
  
ggplotly(p)
```

This just tells us which vehicles are similar to one another by car make. We can view the loadings (coefficients per original variable) for our PCA results:

```{r tidymodels-pca-cluster, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Principal Component Analysis loadings for mtcars

 p <- pca_prep |>
  tidy_clust(5) |>
  filter(component %in% c("PC1", "PC2")) |>
  group_by(component) |>
  slice_max(abs(value), n = 5) |>
  arrange(component, desc(abs(value))) |>
  mutate(absval = abs(value)) |>
  group_by(component) |>
  slice_max(absval, n = 5) |>
  ggplot(aes(x = reorder(terms, value), y = value, fill = component)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~component, scales = "free_y") +
  theme_minimal() +
  labs(x = "Variable", y = "Loading weight")

ggplotly(p)
```

For the first principal component (PC1):

- Displacement (disp), Cylinders (cyl), Weight (wt), and Horsepower (hp) all have positive loadings, meaning they increase together along PC1. Cars with high PC1 scores tend to be heavier, more powerful, and larger-engined.
- Engine shape (vs) has a negative loading, meaning it moves opposite to PC1. Higher PC1 scores are associated with V-shaped engines (vs = 0), while straight engines (vs = 1) correspond to lower PC1 scores.

PC1 primarily captures a "size and power vs engine type" gradient across the cars in the dataset.

## K-Means Clustering 

We will next perform K-means clustering on the first two principal components of the mtcars dataset to group cars based on overall patterns in size, power, and efficiency. By reducing the data with PCA, we captured the main variance in fewer dimensions, and K-means then identifies clusters of similar cars. [Click here to learn more about K-means clustering.](https://www.sciencedirect.com/science/article/abs/pii/S0020025522014633)

```{r tidymodels-kmeans-clustering, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Principal Component Analysis on mtcars
set.seed(100)
kmeans_res <- pca_results |>
  select(-car) |>
  kmeans(centers = 3)
  
p <- pca_results |>
  mutate(cluster = as.factor(kmeans_res$cluster)) |>
  ggplot(aes(PC1, PC2, color = cluster, label = car)) +
  geom_point(size = 1) +
  theme_minimal() +
  labs(x = paste0("PC1 (", pca_explained[1], " variance explained)"), y = paste0("PC2 (", pca_explained[2], " variance explained)"))
  
ggplotly(p)
```

The resulting plot shows possible groupings, with clusters reflecting differences in engine size, weight, and horsepower. Have a play around with different values of k and see how that influences the clustering results.

# Cross Validation

We will next set up k-fold cross validation using k=5 folds, the purpose of which is to reliably estimate a model's performance on unseen data while making the most of the available dataset. [More information on cross-validation can be found here.](https://pmc.ncbi.nlm.nih.gov/articles/PMC11412612/)

Our training set will be divided into 5 segments to assess its predictive performance
and tune the hyper-parameters of the machine learning models. We will train our model on 4 folds and assess it against the remaining validation fold. This process is repeated for each fold.

```{r tidymodels-cv, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
folds <- vfold_cv(
  data_train, v = 5
)
```

For hyper-parameter optimisation, we will combine 5-fold cross
validation with a grid search over different hyper-parameter
combinations. [A comprehensive tutorial on grid-search may be found here.](https://www.tmwr.org/grid-search)

We will aim to optimise the predictive accuracy of the
machine learning models using the root mean square (RMSE) and mean
absolute error (MAE) metrics.

```{r tidymodels-grid-search, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
metric <- metric_set(rmse, mae)

ctrl_grid <- control_grid(
  save_pred = TRUE,
  save_workflow = TRUE,
  extract = identity,
  parallel_over = "resamples",
  allow_par = TRUE
)
```

# Workflow Set

We will next use the [workflowsets
package](https://workflowsets.tidymodels.org/) to easily streamline, manage,
combine, and compare several machine learning workflows for multiple
models. 

We next construct 3 workflows composed of:

-   Linear regression: We will use the *glmnet* package.
-   Random forest: We will use the *ranger* package.
-   Linear support vector machine: We will use the *kernlab* package.

```{r tidymodels-ml-recipes, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
lr_recipe <- linear_reg(mixture = tune(), penalty = tune()) |> 
  set_engine("glmnet") |> 
  set_mode("regression")

rf_recipe <- rand_forest(trees = tune()) |> 
  set_engine("ranger") |> 
  set_mode("regression") 

svm_recipe <- svm_linear(margin = tune()) |> 
  set_engine("kernlab") |> 
  set_mode("regression") 
```

We will next combine these 3 model workflows together into a single
workflowset, while incorporating the RMSE and MAE evaluation metrics,
5-fold cross-validation, and grid search hyper-parameter optimisation.

```{r tidymodels-workflowset, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

training_recipe <- data_train |>
  recipe() |>
  update_role(everything(), new_role = "predictor") |>
  update_role(mpg, new_role = "outcome") |>
  step_zv() |>
  step_nzv() |>
  step_normalize(all_numeric(), -all_outcomes()) |>
  step_impute_knn() |>
  step_rm(car)

model_set <- workflow_set(
  preproc = list(
    standard = training_recipe
  ),
  models = list(
    lr = lr_recipe,
    rf = rf_recipe,
    svm = svm_recipe
  ),
  cross = TRUE
) |>
  option_add(
    metrics = metric,
    control = ctrl_grid,
    resamples = folds
  )
```

# Hyper-parameter optimisation

We will define the grid of hyper-parameter values for performing grid
search. The specification for each of our models is as follows:

-   Linear Regression: Mixure defines the weighting of L1 and L2
    regularisation; a value of mixture = 1 corresponds to a pure LASSO
    model, while mixture = 0 indicates ridge regression. Penalty refers
    to the strength of the regularisation.
-   Random Forest: Trees determines the number of ensemble members, and
    hence the size of the ensemble.
-   Support Vector Machine: Margin refers to the epsilon smoothness
    hyper-parameter. Smaller values produce more support vectors which
    may lead to overfitting, while larger values produce fewer support
    vectors which may lead to underfitting.

```{r tidymodels-hyperopt-grid-spec, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
lr_params <- extract_parameter_set_dials(lr_recipe) |>
  update(
    mixture = mixture(c(0, 1)), penalty = penalty(c(0, 5))
  )

rf_params <- extract_parameter_set_dials(rf_recipe) |>
  update(
    trees = trees(c(1, 250))
  )

svm_params <- extract_parameter_set_dials(svm_recipe) |>
  update(
    margin = svm_margin(c(0, 0.25))
  )
```

We next execute the grid search for hyper-parameter tuning.

```{r tidymodels-hyperopt-grid-run, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

model_set <- model_set |>
  option_add_parameters() |>
  option_add(
    param_info = lr_params,
    id = "standard_lr"
  ) |>
  option_add(
    param_info = rf_params,
    id = "standard_rf"
  ) |>
  option_add(
    param_info = svm_params,
    id = "standard_svm"
  ) 

tuned_model_set <- model_set |>
  workflow_map(
    "tune_grid", 
    seed = 1000
  )

tuned_model_set
```

This produces a set of tuned machine learning models.

## Results

We can visualise the results for the hyper-parameter optimisation
of our 5-fold cross-validation procedure. We can compare the predictive
performance of our 3 machine learning models.

```{r tidymodels-hyperopt-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Hyper-parameter optimisation results via grid search for 5-fold cross validation.
workflow_autoplot(tuned_model_set) |> 
  ggplotly()
```

We see that the random forest models appear to have the greatest
predictive accuracy, with linear regression having the poorest, and
support vector machine sitting in the middle. Let's view a table of metrics for the best models.

```{r tidymodels-best-model-metrics, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
tuned_model_set |>
  rank_results(rank_metric = "rmse") |>
  filter(.metric == "rmse") |>
  slice_head(n=5)
```

We finally retrieve the best fitted model, and evaluate it.

```{r tidymodels-best-model, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
best_model <- fit_best(tuned_model_set)
best_model
```



Let's plot the predictions of the best fitted model:

```{r tidymodels-prediction-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Predicted and observed Miles per Gallon for testing set.
p <- data_test |> 
  mutate(
    predicted_mpg = predict(best_model, new_data = data_test)  |> 
      pluck(".pred")
  ) |>
  ggplot(aes(x = wt, y = mpg, color = "Observed MPG")) +
  geom_point() +
  geom_point(aes(x = wt, y = predicted_mpg, color = "Predicted MPG")) +
  labs(x = "Weight", y = "Miles per Gallon (MPG)", color = "Variable") +
  theme_minimal() 

ggplotly(p)
```

# Stacking ensemble

Rather than selecting only a single best model, we may instead
construct an ensemble composed of all the tuned machine learning models.
[We will make use of the *stacks* package to build a stacking
ensemble.](https://stacks.tidymodels.org/). Under the hood, LASSO and ridge regression
meta-models will be trained upon the combined/stacked predictions of all
the random forest, support vector machine, and linear regression models against Miles per Gallon, which generally
results in superior predictive performance. [More information on stacking can be found here.](https://www.sciencedirect.com/science/article/pii/S0893608005800231)

```{r tidymodels-stacking, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

stack_regressor <- stacks() |>
  add_candidates(tuned_model_set) 

invisible(capture.output(
  stack_regressor <- stack_regressor |>
  blend_predictions(
    penalty =  10^(-10:-1),
    mixture = seq(0, 1, length.out = 10),
    times = 10
  ) |>
  fit_members()
))

stack_regressor
```

Above, we can see the models selected by our regularised linear regression meta-model. Let's plot the stacking process for our ensemble:

```{r tidymodels-stacking-auto-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Ensemble stacking results.
p <- stacks_autoplot(stack_regressor) 

ggplotly(p)
```

Finally, let's take a look at the predictions of the stacking ensemble:

```{r tidymodels-stacking-prediction-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Predicted and observed Miles per Gallon for testing set using stacking ensemble.
p <- data_test |> 
  mutate(
    predicted_mpg = predict(stack_regressor, new_data = data_test)  |> 
      pluck(".pred")
  ) |>
  ggplot(aes(x = wt, y = mpg, color = "Observed MPG")) +
  geom_point() +
  geom_point(aes(x = wt, y = predicted_mpg, color = "Predicted MPG")) +
  labs(x = "Weight", y = "Miles per Gallon (MPG)", color = "Variable") +
  theme_minimal() 

ggplotly(p)
```

# Uncertainty Quantification

Our stacking ensemble does not produce prediction intervals
out-of-the-box. Hence, no in-built uncertainty quantification. There are
various ways of constructing prediction intervals for our ensemble. [We
will use a non-parametric (no distributional assumptions) and
model-agnostic approach via the *probably*
package](https://probably.tidymodels.org/). We will use *probably* to
conformalise the stacking ensemble to produce non-parametric prediction
intervals using conformal inference. [Refer to this document for more
information on conformal inference.](https://arxiv.org/abs/2107.07511)

```{r tidymodels-conformalisation, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
patched_stack_regressor <- tidymodels_deployment$patch_tidymodel(stack_regressor, training_recipe, data_train)
set.seed(200)
conformalised_regressor <- int_conformal_split(patched_stack_regressor, data_train)
conformalised_regressor$training <- data_train
```

We can next view a calibration plot of our conformalised model.

```{r fig-conformal-ensemble-work-calibration, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Calibration plot of observed and predicted Miles per Gallon on testing data.

p <- conformalised_regressor |>
  predict(new_data = data_test, level = 0.95) |>
  mutate(
    mpg = data_test$mpg
  ) |>
  cal_plot_regression(truth = mpg, estimate = .pred) 

ggplotly(p)
```

In this case, a diagonal line means that our model is producing
reasonably well-calibrated prediction intervals. We can view the
predictions produced by the conformalised model, this time with
prediction intervals included:

```{r tidymodels-conformalised-prediction-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Predicted and observed Miles per Gallon for testing set using conformalised model (95% prediction intervals).

p <- conformalised_regressor |> 
  predict(new_data = data_test, level = 0.95) |>
  mutate(
    mpg = data_test$mpg,
    wt = data_test$wt
  ) |>
  ggplot(aes(x = wt, y = mpg, color = "Observed MPG")) +
  geom_point() +
  geom_point(aes(x = wt, y = .pred, color = "Predicted MPG")) +
  geom_ribbon(
    aes(ymin = .pred_lower, ymax = .pred_upper, colour = "Predicted MPG"),
    alpha = 0.2
  ) +
  labs(x = "Weight", y = "Miles per Gallon (MPG)", color = "Variable") +
  theme_minimal() 

ggplotly(p)
```

Most of the testing observations fall within the 95% prediction
intervals. Given a large enough testing dataset, we would expect roughly
5% of new observations to fall outside of the intervals.

# Model Interpretation

A conformalised stacking ensemble (or many types of complex machine learning models) will be relatively black-box in
nature. Ideally, we would want to be able to interrogate and interpret
our model. [The DALEX package can be used to facilitate this
interpretability.](https://dalex.drwhy.ai/) For this tutorial, we will
demonstrate three functions offered by DALEX.

## Feature importances

We will rank the importance of each feature by using a permutation-based
approach. Here, we will evaluate how much a model's prediction error
increases when you randomly shuffle a feature's values. If shuffling a
feature doesn't change the error then the feature wasn't important. If
shuffling a feature makes the error spike then the feature is important
for predictions. [More information can be found here.](https://ema.drwhy.ai/featureImportance.html)

```{r fig-conformal-global-explain, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Permutation-based feature importances for the conformalised model.

conformal_explainer <- 
  explain_tidymodels(
    model = conformalised_regressor,
    data = data_train |> select(-mpg), 
    y = data_train |>  pull(mpg),
    label = "conformal"
  )

p <- feature_importance(
  conformal_explainer,
  type = "variable_importance",
  n_sample = 1000
) |>
  plot()
 
ggplotly(p)
```

The top 4 most important features are:

-   disp (Displacement): Engine displacement in cubic inches.
-   hp (Horsepower): Engine power in brake horsepower (bhp).
-   cyl (Cylinders): Number of cylinders in the engine.
-   wt (Weight): Weight of the car in 1000 lbs.

## Partial Dependence

We can next use a partial dependence plot (PDP) to visualise the effect
of one or more features on Miles per Gallon for our model, while
averaging out the effects of all other features. [Open this link for more information.](https://ema.drwhy.ai/partialDependenceProfiles.html)

```{r fig-conformal-pdp, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Partial dependence plots for the conformalised model.

p <- model_profile(
  conformal_explainer,
  type = "partial"
) |>
  plot() 

ggplotly(p)
```

We see that smaller values of Displacement, Horsepower, and Weight
typically have more Miles per Gallon. There is a similar trend for the
number of Cylinders, with a drop-off at around 7 cylinders.

## Residuals

We will next produce a residuals plot to determine how well the model
predicted Miles per Gallon at different values. [Open this link for more information.](https://ema.drwhy.ai/residualDiagnostic.html)

```{r fig-conformal-residuals, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Residual plot for the conformalised model.

p <- model_diagnostics(
  conformal_explainer
) |>
  plot(variable = "y_hat", yvariable = "residuals")

ggplotly(p)
```

We see an S-shaped curve, suggesting that there is bias in the
predictions. There is evidence of over-prediction for Miles per Gallon
below 12.5 and between 20 and 29. Conversely, there is evidence of
under-prediction for Miles per Gallon over 29 and between 12.5 and 20.

# Model versioning, sharing, and deployment

Finally, we will briefly demonstrate how to perform model versioning, sharing, and
deployment using a combination of:

-   [pins: For sharing and versioning resources and
    artifacts](https://pins.rstudio.com/)
-   [vetiver: For model versioning, sharing, and
    deployment](https://rstudio.github.io/vetiver-r/)
-   [plumber: For creating light-weight web applications, and deploying models
    as microservices](https://www.rplumber.io/)

In this example, we will first save our conformalised model to a local temporary directory using *pins* and *vetiver*

```{r tidymodels-vetiver-save, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
board <- board_temp()
v_model <- vetiver_model(conformalised_regressor, "conformal_ensemble")
vetiver_pin_write(v_model, board = board)
```

We will next load and interact with the saved model using a plumber web application.

```{r tidymodels-vetiver-load, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
loaded_v <- board |> 
  vetiver_pin_read("conformal_ensemble")

pr(here("app", "api", "plumber.R")) |>
  vetiver_api(loaded_v) |>
  pr_run(port = 8088)
```

# Model cards

[A model card](https://doi.org/10.1145/3287560.3287596) is a
standardised document that provides key information about a machine
learning model. It is essentially a "spec sheet" for an AI model. It’s
meant to help people understand what the model does, how it was trained,
and its limitations. *vetiver* offers support for generating model cards
from an RMD template file. Run the below code to view an example
template (don't try and render the template).

```{r tidymodels-model-card, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
file.edit(here("pages", "tidymodels_workshop", "cards", "card.rmd"))
```

We have already rendered the model card as an HTML file. This HTML file
can be saved and retrieved using *pins*:

```{r tidymodels-card-save, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
board |> 
  pin_upload(
    here("pages", "tidymodels_workshop", "cards", "card.html"), 
    "conformal_ensemble_card"
  )
```

We will finally load and interact with the saved model using a plumber
web application. This time, with a model card available.

```{r tidymodels-vetiver-card-load, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
Sys.setenv("MODEL_CARD_NAME" = "conformal_ensemble_card")

loaded_v <- board |> 
  vetiver_pin_read("conformal_ensemble")

pr(here("app", "api", "plumber.R")) |>
  vetiver_api(loaded_v) |>
  pr_run(port = 8088)
```

# Bonus: Pipelines with targets

[The *targets* package](https://books.ropensci.org/targets/) is used to
construct reproducible data/statistical science workflows in R. A
workflow is composed of a sequence of steps which are executed after the
other. Successfully executed steps are cached to obviate the need to
re-run already completed code.

[*targets* can be combined with *crew* to facilitate distributed and
parallelised computing within different execution
environments.](https://books.ropensci.org/targets/crew.html)

To view the targets pipeline code, run the following:

```{r tidymodels-targets-pipeline, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
file.edit(here("pipelines", "_targets_tidymodels_deployment.R"))
```

Next, we can run the targets pipeline.

```{r tidymodels-targets, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}

tar_make(
  script = here("pipelines", "_targets_tidymodels_deployment.R"),
  use_crew = TRUE
)
```

