---
title: "Machine Learning Workflow using H2O"
date: today
authors: "James Bristow, Jingjing Zhang, Hana Liang, Lindy Guo"
fig-pos: 'H'
toc: true
lof: true
lot: true
link-citations: true
editor: 
  markdown: 
    wrap: 72
execute:
  warning: false
  message: false
---

# Introduction

In this section, we will demonstrate how one may train a machine
learning model and tune its hyper-parameters using the *tidymodels*
framework. [For more information, click this
link.](https://www.tidymodels.org/)

[We will combine *tidymodels* with H2O.](https://h2o.ai/) H2O is an
open-source, distributed machine learning platform designed for scalable
and fast data analysis. It’s widely used for building predictive models
on large datasets and supports both traditional machine learning
algorithms and modern AI methods.

We first load all required dependencies.

```{r setup, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, output=TRUE, eval=TRUE}
box::use(../../R/proj_utils)
box::use(../../R/tidymodels_deployment)

box::use(
  dials[svm_margin, trees, penalty, mixture],
  readr[read_csv, read_file],
  tune[control_grid, extract_parameter_set_dials, .get_tune_metric_names, select_best],
  here[here],
  dplyr[mutate, bind_rows, arrange, filter, select, pull], 
  purrr[pluck],
  ranger[ranger],
  kernlab,
  rsample[initial_split, vfold_cv, training, testing],
  parsnip[rand_forest, svm_linear, boost_tree, set_engine, set_mode, linear_reg, auto_ml],
  recipes[recipe, update_role, step_zv, step_nzv, step_normalize, step_impute_knn, update, all_numeric, all_outcomes],
  yardstick[rmse, mae, metric_set, metrics],
  hardhat[tune],
  ggplot2[ggplot, geom_point, labs, theme_minimal, aes, geom_ribbon, geom_boxplot, scale_x_sqrt],
  workflows[workflow, add_recipe, add_model, fit],
  workflowsets[workflow_set, option_add, option_add_parameters, workflow_map, fit_best, autoplot],
  stacks[stacks, add_candidates, blend_predictions, fit_members],
  vetiver[vetiver_model, vetiver_pin_write, vetiver_api, vetiver_pin_read],
  pins[board_temp, board_local, board_s3, pin_write, pin_read, pin_upload, pin_download],
  paws.common,
  paws.storage,
  plotly[ggplotly],
  probably[int_conformal_split, cal_plot_regression],
  plumber[pr, pr_run],
  DALEX[feature_importance, model_profile, predict_profile, model_diagnostics],
  DALEXtra[explain_tidymodels],
  glmnet,
  zeallot[...],
  targets[tar_make],
  h2o[h2o.connect, h2o.init],
  agua[extract_fit_parsnip, rank_results, member_weights],
  tidyr[unnest]
)

source(here("R/probably_vetiver.R"))
```

# Data

We will next load an example dataset. [For this tutorial, we will use
the mtcars
dataset.](https://cran.r-project.org/web/packages/explore/vignettes/explore-mtcars.html)

```{r load-h2o-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

data("mtcars")
head(mtcars)
```

We will next perform a 80-20 train-test data split. We will train our
model on 80% of the data, and assess it against the other 20%.

```{r split-h2o-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
data_split <- mtcars |>
  initial_split(
    prop = 0.8
  )

data_train <- training(data_split)
data_test  <- testing(data_split)

head(data_train)
```

Next, we will create a data preprocessing recipe using the [recipes
package](https://recipes.tidymodels.org/). This recipe will:

-   Drop zero and near zero-variance columns.
-   Normalise all numeric columns using Z-standardisation.
-   Impute missing data using k nearest neighbor imputation.

```{r process-h2o-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
training_recipe <- recipe(data_train) |>
  update_role(everything(), new_role = "predictor") |>
  update_role(mpg, new_role = "outcome") |>
  step_zv() |>
  step_nzv() |>
  step_normalize(all_numeric(), -all_outcomes()) |>
  step_impute_knn()
```

# AutoML

Rather than using *workflowsets* and various R-based models, we will
instead launch an H2O server and execute its supported AutoML algorithm.
We will only run this algorithm for a maximum of 20 seconds for
demonstration purposes. You would typically want to run it for longer
depending on the nature of the dataset.

```{r h2o-automl, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
h2o.init(max_mem_size = "512M", nthreads = 1)

auto_spec <-
  auto_ml() |>
  set_engine("h2o", max_runtime_secs = 20, seed = 1) |>
  set_mode("regression")

auto_wflow <-
  workflow() |>
  add_model(auto_spec) |>
  add_recipe(training_recipe)

auto_fit <- fit(auto_wflow, data = data_train)
extract_fit_parsnip(auto_fit)
```

## Results

We next view the evaluation results of the AutoML algorithm, sorting by
predictive accuracy using the mean absolute error (MAE) metric.

```{r h2o-automl-results, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

rank_results(auto_fit) |>
  filter(.metric == "mae") |>
  arrange(rank)
```

The generalised linear models (GLMs) and gradient boosting machine (GBM)
algorithms show promising results. We can next visualise the stacking
ensemble member weights from H2O's AutoML algorithm.

```{r fig-h2o-importances, results = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: H20 AutoML model importances.

p <- auto_fit |>
  extract_fit_parsnip() |>
  member_weights() |>
  unnest(importance) |>
  filter(type == "scaled_importance") |>
  ggplot() +
  geom_boxplot(aes(value, algorithm)) +
  scale_x_sqrt() +
  labs(y = NULL, x = "scaled importance", title = "Member importance in AutoML model")

ggplotly(p)
```

This plot shows that the GLM and GBM models are more heavily weighted in
the ensemble than the other supported algorithms.

# Uncertainty Quantification

Our AutoML model does not produce prediction intervals out-of-the-box.
Hence, no in-built uncertainty quantification. We will use *probably* to
conformalise the model to produce non-parametric prediction intervals
using conformal inference.

```{r h2o-conformalisation, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
patched_stack_regressor <- tidymodels_deployment$patch_tidymodel(auto_fit, training_recipe, data_train)
conformalised_regressor <- int_conformal_split(patched_stack_regressor, data_train)
conformalised_regressor$training <- data_train
```

We can view the predictions produced by the conformalised model, this
time with prediction intervals included:

```{r fig-h2o-conformalised-prediction-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Predicted and observed Miles per Gallon for testing set using conformalised model (95% prediction intervals).

p <- conformalised_regressor |> 
  predict(new_data = data_test, level = 0.95) |>
  mutate(
    mpg = data_test$mpg,
    wt = data_test$wt
  ) |>
  ggplot(aes(x = wt, y = mpg, color = "Observed MPG")) +
  geom_point() +
  geom_point(aes(x = wt, y = .pred, color = "Predicted MPG")) +
  geom_ribbon(
    aes(ymin = .pred_lower, ymax = .pred_upper, colour = "Predicted MPG"),
    alpha = 0.2
  ) +
  labs(x = "Weight", y = "Miles per Gallon (MPG)", color = "Variable") +
  theme_minimal() 

ggplotly(p)
```

Most of the testing observations fall within the 95% prediction
intervals. Given a large enough testing dataset, we would expect roughly
5% of new observations to fall outside of the intervals.

# Model Interpretation

An AutoML model will be relatively black-box in nature. For this
tutorial, we will demonstrate three functions offered by DALEX.

## Partial Dependence

We can next use a partial dependence plot (PDP) to visualise the effect
of one or more features on Miles per Gallon for our model, while
averaging out the effects of all other features.

```{r fig-conformal-pdp, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Partial dependence plots for the AutoML model.

conformal_explainer <- 
  explain_tidymodels(
    model = conformalised_regressor,
    data = data_train |> select(-mpg), 
    y = data_train |>  pull(mpg),
    label = "conformal"
  )

p <- model_profile(
  conformal_explainer,
  type = "partial"
) |>
  plot() 

ggplotly(p)
```

We see that smaller values of Displacement, Horsepower, and Weight
typically have more Miles per Gallon. There is a similar trend for the
number of Cylinders, with a drop-off at around 5 and 7 cylinders.

## Residuals

We will next produce a residuals plot to determine how well the model
predicted Miles per Gallon at different values.

```{r fig-conformal-residuals, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Residual plot for the AutoML model.

p <- model_diagnostics(
  conformal_explainer
) |>
  plot(variable = "y_hat", yvariable = "residuals")

ggplotly(p)
```

We see an S-shaped curve, suggesting that there is bias in the
predictions. There is evidence of over-prediction for Miles per Gallon
between 10 and 17, and between 21 and 28. Conversely, there is evidence
of under-prediction for Miles per Gallon over 28 and between 17 and 22.

# Model versioning, sharing, and deployment

Finally, we will demonstrate how to perform model versioning, sharing,
and deployment using a combination of *pins*, *vetiver*, and *plumber*.
In this example, we will first save our AutoML model to a local
temporary directory using *pins* and *vetiver*

```{r h2o-vetiver-save, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
board <- board_temp()
v_model <- vetiver_model(conformalised_regressor, "automl_model")
vetiver_pin_write(v_model, board = board)
```

We will next load and interact with the saved model using a plumber web
application.

```{r h2o-vetiver-load, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
loaded_v <- board |> 
  vetiver_pin_read("automl_model")

pr(here("app", "api", "plumber.R")) |>
  vetiver_api(loaded_v) |>
  pr_run(port = 8088)
```

# Bonus: Pipelines with targets

[The *targets* package](https://books.ropensci.org/targets/) is used to
construct reproducible data/statistical science workflows in R.

To view the targets pipeline code, run the following:

```{r h2o-targets-pipeline, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
file.edit(here("pipelines", "_targets_h2o_deployment.R"))
```

Next, we can run the targets pipeline.

```{r h2o-targets, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}

tar_make(
  script = here("pipelines", "_targets_h2o_deployment.R"),
  use_crew = TRUE
)
```
