---
title: "MLOps using vetiver"
date: today
authors: "James Bristow, Jingjing Zhang, Hana Liang, Lindy Guo"
fig-pos: 'H'
toc: true
lof: true
lot: true
link-citations: true
editor: 
  markdown: 
    wrap: 72
execute:
  warning: false
  message: false
---

# Introduction

In this section, we will demonstrate how one may combine *vetiver* and
*plumber* with *Docker* and *Kubernetes* to enable Machine Learning
Operations (MLOps) practices. MLOps is the practice of combining machine
learning (ML) development with software engineering and operational
practices. It applies principles from DevOps like continuous
integration, continuous delivery (CI/CD), version control, and
monitoring to the lifecycle of ML models to make ML systems
reproducible, scalable, and reliable. Models can then be updated safely,
deployed efficiently, and monitored for performance drift over time

As a reminder:

-   [vetiver: For model versioning, sharing, and
    deployment](https://rstudio.github.io/vetiver-r/)
-   [plumber: For creating light-weight web applications, and deploying
    models as microservices](https://www.rplumber.io/)
-   [pins: For uploading, downloading, and sharing
    resources.](https://pins.rstudio.com/)

We will elucidate upon *Docker* and *Kubernetes* in subsequent sections.

We first load all required dependencies.

```{r setup, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, output=TRUE, eval=TRUE}
if (Sys.getenv("S3_ENDPOINT_URL") == "") {
  Sys.setenv("S3_ENDPOINT_URL" = "http://127.0.0.1:9000")
}

box::use(../../R/proj_utils)
box::use(../../R/tidymodels_deployment)

box::use(
  dials[svm_margin, trees, penalty, mixture],
  readr[read_csv, read_file],
  tune[control_grid, extract_parameter_set_dials, .get_tune_metric_names, select_best],
  here[here],
  dplyr[mutate, bind_rows, arrange, select, pull],
  purrr[pluck],
  ranger[ranger],
  kernlab,
  rsample[initial_split, vfold_cv, training, testing],
  parsnip[rand_forest, svm_linear, boost_tree, set_engine, set_mode, linear_reg],
  recipes[recipe, update_role, step_zv, step_nzv, step_normalize, step_impute_knn, update, all_numeric, all_outcomes],
  yardstick[rmse, mae, metric_set, metrics],
  hardhat[tune], 
  htmlwidgets[saveWidget],
  ggplot2[ggplot, geom_point, labs, theme_minimal, aes, geom_ribbon],
  workflows[workflow, add_recipe, add_model, fit],
  workflowsets[workflow_set, option_add, option_add_parameters, workflow_map, fit_best, workflow_autoplot = autoplot],
  stacks[stacks, add_candidates, blend_predictions, fit_members],
  vetiver[vetiver_model, vetiver_pin_write, vetiver_api, vetiver_pin_read],
  pins[board_temp, board_local, board_s3, pin_write, pin_read, pin_upload, pin_download],
  paws.common,
  paws.storage,
  plotly[ggplotly],
  probably[int_conformal_split, cal_plot_regression],
  plumber[pr, pr_run],
  DALEX[feature_importance, model_profile, predict_profile, model_diagnostics],
  DALEXtra[explain_tidymodels],
  glmnet,
  zeallot[...],
  targets[tar_make]
)

source(here("R/probably_vetiver.R"))
```

# Data

We will next load an example dataset. [For this tutorial, we will use
the mtcars
dataset.](https://cran.r-project.org/web/packages/explore/vignettes/explore-mtcars.html)

```{r load-tidymodels-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

data("mtcars")
head(mtcars)
```

We will next perform a 80-20 train-test data split.

```{r split-tidymodels-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
data_split <- mtcars |>
  initial_split(
    prop = 0.8
  )

data_train <- training(data_split)
data_test  <- testing(data_split)

head(data_train)
```

Next, we will create a data preprocessing recipe using the [recipes
package](https://recipes.tidymodels.org/), which will:

-   Drop zero and near zero-variance columns.
-   Normalise all numeric columns using Z-standardisation.
-   Impute missing data using k nearest neighbor imputation.

```{r process-tidymodels-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
training_recipe <- recipe(data_train) |>
  update_role(everything(), new_role = "predictor") |>
  update_role(mpg, new_role = "outcome") |>
  step_zv() |>
  step_nzv() |>
  step_normalize(all_numeric(), -all_outcomes()) |>
  step_impute_knn()
```

# Training

We will train a simple random forest model with default settings using
tidymodels:

```{r tidymodels-ml-rf, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
rf_recipe <- rand_forest() |> 
  set_engine("ranger") |> 
  set_mode("regression") 

rf_workflow <- workflow() |> 
  add_recipe(training_recipe) |> 
  add_model(rf_recipe)

fit_rf <- fit(rf_workflow, data = data_train) 
```

# Uncertainty Quantification

Most implementations of random forest do not produce prediction
intervals out-of-the-box. [We will use a non-parametric (no
distributional assumptions) and model-agnostic approach to constructing
prediction intervals via the *probably*
package](https://probably.tidymodels.org/).

```{r tidymodels-conformalisation, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
conformalised_rf <- int_conformal_split(fit_rf, data_train)
conformalised_rf$training <- data_train
```

We can view the predictions of the random forest model:

```{r fig-tidymodels-conformalised-rf-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| fig-cap: Predicted and observed Miles per Gallon for testing set using conformalised random forest (95% prediction intervals).

p <- conformalised_rf |> 
  predict(new_data = data_test, level = 0.95) |>
  mutate(
    mpg = data_test$mpg,
    wt = data_test$wt
  ) |>
  ggplot(aes(x = wt, y = mpg, color = "Observed MPG")) +
  geom_point() +
  geom_point(aes(x = wt, y = .pred, color = "Predicted MPG")) +
  geom_ribbon(
    aes(ymin = .pred_lower, ymax = .pred_upper, colour = "Predicted MPG"),
    alpha = 0.2
  ) +
  labs(x = "Weight", y = "Miles per Gallon (MPG)", color = "Variable") +
  theme_minimal() 

ggplotly(p)
```

Most of the testing observations fall within the 95% prediction
intervals, though the predicted Miles per Gallon for larger weights are
quite poor.

# Model versioning, sharing, and deployment

We will demonstrate how to perform model versioning, sharing, and
deployment using a combination of *pins*, *vetiver*, and *plumber*.
[We'll store everything in S3. Amazon S3 (Simple Storage Service) is
cloud storage for files, called "objects", organised in "buckets". It
offers scalable, secure, and flexible storage with features like
versioning, encryption, and programmatic access via URLs or AWS
tools.](https://docs.aws.amazon.com/AmazonS3/latest/userguide/Welcome.html)
In this example, we will first specify our S3 bucket.

```{r tidymodels-s3-board, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
board <- board_s3(
  "data", 
  access_key = "user", 
  secret_access_key = "password", 
  region = "us-east-2", 
  endpoint = Sys.getenv("S3_ENDPOINT_URL") 
)
```

We will generate and save several plots for our random forest model to
upload into an S3 bucket:

::: {.callout-tip title="Plots" collapse="true"}
```{r tidymodels-vetiver-plots-save, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
save_s3_plot <- function(p, object_name, model, board) {
  outfile <- paste0(object_name, ".html")

  p |> 
    ggplotly() |>
    saveWidget(file.path("/tmp", outfile), selfcontained = TRUE)
    
  model <- proj_utils$model_s3_upload(
    model, board, Sys.getenv("S3_ENDPOINT_URL"), 
    "/tmp", outfile, object_name
  )
  model
}

conformalised_rf <- conformalised_rf |>
  predict(new_data = data_test, level = 0.95) |>
  mutate(mpg = data_test$mpg) |>
  cal_plot_regression(truth = mpg, estimate = .pred) |>
  save_s3_plot("conformal_rf_calibration", conformalised_rf, board)

conformal_explainer <- 
  explain_tidymodels(
    model = conformalised_rf,
    data = data_train |> select(-mpg), 
    y = data_train |>  pull(mpg),
    label = "conformal"
  )

conformalised_rf <- feature_importance(
  conformal_explainer,
  type = "variable_importance",
  n_sample = 1000
) |>
  plot() |> 
  save_s3_plot("conformal_rf_variable_importance", conformalised_rf, board)

conformalised_rf <- model_profile(
  conformal_explainer,
  type = "partial"
) |>
  plot() |> 
  save_s3_plot("conformal_rf_pdp", conformalised_rf, board)

conformalised_rf <- model_diagnostics(
  conformal_explainer
) |>
  plot(variable = "y_hat", yvariable = "residuals") |> 
  save_s3_plot("conformal_rf_residuals", conformalised_rf, board)
```
:::

*vetiver* offers support for generating model cards from an RMD template
file. We will store a rendered model card as an HTML file within an S3
bucket using *pins*:

```{r tidymodels-card-save, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
conformalised_rf <- proj_utils$model_s3_upload(
  conformalised_rf, 
  board, 
  Sys.getenv("S3_ENDPOINT_URL"), 
  here("pages", "tidymodels_workshop", "cards"), 
  "card.html", 
  "random_forest_card",
  is_card = TRUE
)
```

We will save our random forest model to an S3 bucket using *pins* and
*vetiver*.

```{r tidymodels-vetiver-save, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
v_model <- vetiver_model(conformalised_rf, "random_forest")
vetiver_pin_write(v_model, board = board)
```

We will next load and interact with the saved model locally using a
plumber web application.

```{r tidymodels-vetiver-load, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
Sys.setenv("MODEL_CARD_NAME" = "random_forest_card")

loaded_v <- board |> 
  vetiver_pin_read("random_forest")

pr(here("app", "api", "plumber.R")) |>
  vetiver_api(loaded_v) |>
  pr_run(port = 8088)
```

# Docker

[Docker is a containerisation technology that allows developers to
package an application along with all its dependencies, libraries, and
configuration files into a single unit called a
container](https://www.docker.com/). This ensures that the application
runs consistently across different environments, whether on a
developer's laptop, a testing server, or a production cloud environment.
Docker helps developers "build once, run anywhere" without worrying
about environment mismatches.

Our objective is to package up our plumber web application and model
within a Docker container for subsequent deployment. In brief:

1.  [We have a Dockerfile which contains a sequence of steps needed to
    build a Docker image. We copy our code from the R workshop repo into
    the Docker image (including the plumber web application
    code).](https://github.com/Plant-Food-Research-Open/r-pipeline-development-workshop/blob/main/dockerfiles/shiny/Dockerfile.0.1.0)
2.  [We define a pipeline that builds the Docker image and pushes it
    into a container
    repository.](https://github.com/Plant-Food-Research-Open/r-pipeline-development-workshop/blob/main/.github/workflows/build-r.yaml)
3.  [Our stored image may be found in the GitHub container
    repository.](https://github.com/Plant-Food-Research-Open/r-pipeline-development-workshop/pkgs/container/shiny-rpipelines)
4.  We can then pull down our stored image and instantiate a container
    to deploy our plumber web application, either using Docker directly
    or with a container orchestration platform such as Kubernetes.
5.  Our model can then be loaded into the web application within the
    container via S3.

# Kubernetes

[Kubernetes is an open-source platform that automates the deployment,
scaling, and management of containerised applications, such as our
deployed models.](https://kubernetes.io/)

It organizes containers into logical units called pods, schedules them
across a cluster of machines, and handles networking, storage, and load
balancing. This ensures that your applications run reliably even as
demand changes or failures occur. Kubernetes also provides self-healing
features, automatically restarting or rescheduling containers when they
fail.

In brief:

1.  Kubernetes will pull down our Docker image containing our plumber
    web application.
2.  Kubernetes will then deploy our web application within a pod inside
    our Kubernetes cluster.
3.  Our S3 server is also running on Kubernetes, allowing us to load our
    model and any other artifacts of interest.

The following code will apply our Kubernetes manifest files, which will
subsequently deploy our models:

```{r tidymodels-k8s-apply, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
system("kubectl delete httproute r-pipelines-models")
system("kubectl delete services r-pipelines-models")
system("kubectl delete deployments r-pipelines-models")

system("kubectl apply -f /home/rstudio/deployments/k8s/nzsa")
```
