---
title: "Causal Inference and directed acyclic graphs"
date: today
authors: "Tom Moore, James Bristow"
fig-pos: 'H'
toc: true
lof: true
lot: true
link-citations: true
editor: 
  markdown: 
    wrap: 72
execute:
  warning: false
  message: false
---

```{r setup, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, output=TRUE, eval=TRUE}
suppressWarnings(
  suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(readxl)
  library(here)
  library(janitor)
  library(marginaleffects)
  library(pins)
  library(reshape2)
  library(arrow)
  library(cowplot)
  library(modelr)
  library(bnlearn)
  library(future)
  library(dagitty)
  library(cmdstanr)
  library(brms)
  library(ggdag)
  library(tidybayes)
  library(ggtext)
  library(ggplot2)
  library(simDAG)
  })
)

bayesplot::color_scheme_set("blue")                                   
options(mc.cores = 1)                           
```

# Introduction

In this section, we will cover causal inference via Stan and tools such as daggity.

```{r dag-basic, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
basic_dag <- dagify(
  Y ~ X + A,
  X ~ A,
  B ~ X + Y,
  exposure = "X",
  outcome = "Y",
  coords = list(
    x = c(X = 1, Y = 3, A = 2, B = 2),
    y = c(X = 1, Y = 1, A = 2, B = 0)
  )
)

basic_dag_plot <- basic_dag |> 
  tidy_dagitty() |> 
  mutate(var_type = case_when(
    name == "X" ~ "Exposure",
    name == "Y" ~ "Outcome",
    str_detect(name, "A") ~ "Confounder",
    str_detect(name, "B") ~ "Collider"
  )) |> 
  mutate(var_label = case_match(name,
    "X" ~ "Treatment",
    "Y" ~ "Outcome",
    "A" ~ "",
    "B" ~ ""
  ))

p1 <- ggplot(basic_dag_plot, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(aes(color = var_type), size = 12) +
  geom_richtext(
    aes(label = name), fontface = "bold", color = "white",
    fill = NA, label.color = NA,
    label.padding = grid::unit(c(3, 0, 0, 0), "pt")
  ) + 
  geom_dag_text(
    data = filter(basic_dag_plot, var_type != "Confounder"), aes(label = var_label), 
    nudge_y = -0.25, color = "black", size = 11, size.unit = "pt"
  ) +
  geom_dag_text(
    data = filter(basic_dag_plot, var_type == "Confounder"), aes(label = var_label), 
    nudge_y = 0.26, color = "black", size = 11, size.unit = "pt"
  ) +
   geom_text(
    aes(x = (x + xend) / 2, y = (y + yend + 0.4) / 2, label =  c("","1.10","","","-1.70","-1.90")),
    color = "black", size = 5
  ) +
  scale_color_manual(values = c("#44AA99", "#6699CC", "#882255","#999933"), guide = "none") +
  theme_dag()
```


::: {.callout-tip title="Simulated data" collapse="true"}
```{r dag-sim, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| label: Simulated data
#| message: false

dag <- empty_dag()

dag <- dag + 
  # parent nodes
  node("a", type="rnorm", mean = 10, sd=4) +
  
  # child nodes
  node("x", type="gaussian", parents=c("a"),     betas=c(3.6),
       intercept=12,  error = 2) +
  node("y", type="gaussian", parents=c("a","x"), betas=c(1.1, -1.7),
       intercept=12,  error = 2) +
  node("b", type="gaussian", parents=c("x","y"), betas=c(2.5, -1.9),
       intercept=-15, error = 2) 

set.seed(42)
sim_dat <- sim_from_dag(dag=dag, n_sim=1000)

head(sim_dat)
```
:::

```{r dag-plot, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| out-width: 100%
#| fig-height: 4
#| fig-width: 6
#| warning: false
#| echo: false

p1
```

**Figure 1:** DAG of the simulated data that shows the relationships (arrows) between the outcome (Y), treatment (X), and other measured variables (A + B). The numbers on the arrows represent the known magnitude and direction of relationships with Y.

::: {.callout-tip title="Under the hood" collapse="true"}
```{r dag-vis, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| out-width: 50%
#| fig-height: 5
#| fig-width: 6
#| echo: false
#| warning: false
#| fig-show: hold
#| results: false
#| layout-ncol: 2

sim_dat |>
ggplot(aes(y = y, x = x, color = a)) +
  theme_classic() +
  geom_jitter(width = 10, height = 10) +
  geom_smooth(color = "black", method = "lm", formula = 'y ~ x') +
  scale_color_viridis_c() +
  xlim(-20,120) +
  ylim(-150,20)
  
sim_dat |>
ggplot(aes(y = y, x = x, color = b)) +
  theme_classic() +
  geom_jitter(width = 10, height = 10) +
  geom_smooth(color = "black", method = "lm", formula = 'y ~ x') +
  scale_color_viridis_c() +
  xlim(-20,120) +
  ylim(-150,20)

sim_dat |>
ggplot(aes(y = x, x = a, color = b)) +
  theme_classic() +
  geom_jitter(width = 10, height = 10) +
  geom_smooth(color = "black", method = "lm", formula = 'y ~ x') +
  scale_color_viridis_c()

sim_dat |>
ggplot(aes(y = y, x = a, color = b)) +
  theme_classic() +
  geom_jitter(width = 10, height = 10) +
  geom_smooth(color = "black", method = "lm", formula = 'y ~ x') +
  scale_color_viridis_c()

sim_dat |>
ggplot(aes(y = b, x = x, color = a)) +
  theme_classic() +
  geom_jitter(width = 10, height = 10) +
  geom_smooth(color = "black", method = "lm", formula = 'y ~ x') +
  scale_color_viridis_c()

sim_dat |>
ggplot(aes(y = b, x = y, color = a)) +
  theme_classic() +
  geom_jitter(width = 10, height = 10) +
  geom_smooth(color = "black", method = "lm", formula = 'y ~ x') +
  scale_color_viridis_c()
```
:::

# Science before statistics

- Most approaches fail to recover the simulated value of -1.70.

::: {.panel-tabset}

## Models

```{r dag-brms-model, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| label: models
#| echo: true

fit_brm <- function(formula, model_name) {
  brm(formula,
      family = gaussian(link = "identity"),
      data = sim_dat,
      backend = "cmdstanr",
      file = here("pages", "stan_workshop", "model", model_name),
      threads = threading(3),
      silent = 0,
      refresh = 500,
      chains = 4,
      cores = 4,
      warmup = 2000,
      iter = 4000)
}


# Models
mod_1 <- fit_brm(y ~ x, "mod_1")         # Univariate
mod_2 <- fit_brm(y ~ a, "mod_2")         # Univariate
mod_3 <- fit_brm(y ~ b, "mod_3")         # Univariate
mod_4 <- fit_brm(y ~ x + a + b, "mod_4") # Casual Salad
mod_5 <- fit_brm(y ~ x + a, "mod_5")     # Confounder
mod_6 <- fit_brm(y ~ x + b, "mod_6")     # Collider
```

## Output

```{r dag-brms-model-summary, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| label: model-output
#| echo: false
#| 
summary(mod_1)
summary(mod_2)
summary(mod_3)
summary(mod_4)
summary(mod_5)
summary(mod_6)
```

## Validation

```{r dag-brms-model-validation, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| label: validation
#| echo: false
#| message: false
#| fig-show: hold
#| results: false
#| layout-ncol: 2
#| fig.width: 4
#| fig.height: 2

pp_check(mod_1)
pp_check(mod_2)
pp_check(mod_3)
pp_check(mod_4)
pp_check(mod_5)
pp_check(mod_6)
```

## Coefficients

```{r dag-brms-model-coefficients, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| label: coefficients
#| echo: false

paste("mod_1:", round(summary(mod_1)$fixed$Estimate[[2]], 2))
paste("mod_2:", round(summary(mod_2)$fixed$Estimate[[2]], 2))
paste("mod_3:", round(summary(mod_3)$fixed$Estimate[[2]], 2))
paste("mod_4: ", paste(round(summary(mod_4)$fixed$Estimate[2:4], 2), collapse = ", "))
paste("mod_5:", round(summary(mod_5)$fixed$Estimate[[2]], 2))
paste("mod_6:", round(summary(mod_6)$fixed$Estimate[[2]], 2))
```

## Stan code

```{r dag-brms-model-stan-code, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| label: stan_1
#| echo: false

print("Example Stan code for mod_1")
make_stancode(y ~ x, data = sim_dat, family = gaussian(link = "identity"))
```

:::

```{r dag-conclusion, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| out-width: 50%
#| fig-height: 5
#| fig-width: 6
#| echo: false
#| warning: false
#| results: false
#| layout-ncol: 2

basic_dag <- dagify(
  Y ~ X + A,
  X ~ A,
  B ~ X + Y,
  exposure = "X",
  outcome = "Y",
  coords = list(
    x = c(X = 1, Y = 3, A = 2, B = 2),
    y = c(X = 1, Y = 1, A = 2, B = 0)
  )
)

basic_dag_plot <- basic_dag |> 
  tidy_dagitty() |> 
  mutate(var_type = case_when(
    name == "X" ~ "Exposure",
    name == "Y" ~ "Outcome",
    str_detect(name, "A") ~ "Confounder",
    str_detect(name, "B") ~ "Collider"
  )) |> 
  mutate(var_label = case_match(name,
    "X" ~ "Treatment",
    "Y" ~ "Outcome",
    "A" ~ "",
    "B" ~ ""
  ))

ggplot(basic_dag_plot, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(aes(color = var_type), size = 12) +
  geom_richtext(
    aes(label = name), fontface = "bold", color = "white",
    fill = NA, label.color = NA,
    label.padding = grid::unit(c(3, 0, 0, 0), "pt")
  ) + 
  geom_dag_text(
    data = filter(basic_dag_plot, var_type != "Confounder"), aes(label = var_label), 
    nudge_y = -0.25, color = "black", size = 11, size.unit = "pt"
  ) +
  geom_dag_text(
    data = filter(basic_dag_plot, var_type == "Confounder"), aes(label = var_label), 
    nudge_y = 0.26, color = "black", size = 11, size.unit = "pt"
  ) +
   geom_text(
    aes(x = (x + xend) / 2, y = (y + yend + 0.4) / 2, label =  c("","-5.04","","","-1.40","-0.27")),
    color = "black", size = 5
  ) +
   ggtitle("Univariate \ny ~ x \ny ~ a \ny ~ b") +
  scale_color_manual(values = c("#44AA99", "#6699CC", "#882255","#999933"), guide = "none") +
  theme_dag()

basic_dag <- dagify(
  Y ~ X + A,
  X ~ A,
  B ~ X + Y,
  exposure = "X",
  outcome = "Y",
  coords = list(
    x = c(X = 1, Y = 3, A = 2, B = 2),
    y = c(X = 1, Y = 1, A = 2, B = 0)
  )
)

basic_dag_plot <- basic_dag |> 
  tidy_dagitty() |> 
  mutate(var_type = case_when(
    name == "X" ~ "Exposure",
    name == "Y" ~ "Outcome",
    str_detect(name, "A") ~ "Confounder",
    str_detect(name, "B") ~ "Collider"
  )) |> 
  mutate(var_label = case_match(name,
    "X" ~ "Treatment",
    "Y" ~ "Outcome",
    "A" ~ "",
    "B" ~ ""
  ))

ggplot(basic_dag_plot, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(aes(color = var_type), size = 12) +
  geom_richtext(
    aes(label = name), fontface = "bold", color = "white",
    fill = NA, label.color = NA,
    label.padding = grid::unit(c(3, 0, 0, 0), "pt")
  ) + 
  geom_dag_text(
    data = filter(basic_dag_plot, var_type != "Confounder"), aes(label = var_label), 
    nudge_y = -0.25, color = "black", size = 11, size.unit = "pt"
  ) +
  geom_dag_text(
    data = filter(basic_dag_plot, var_type == "Confounder"), aes(label = var_label), 
    nudge_y = 0.26, color = "black", size = 11, size.unit = "pt"
  ) +
   geom_text(
    aes(x = (x + xend) / 2, y = (y + yend + 0.4) / 2, label =  c("","0.15","","","0.71","-0.42")),
    color = "black", size = 5
  ) +
  ggtitle("Multiple linear regression\n\"Causal salad\" \ny ~ x + a + b\n") +
  scale_color_manual(values = c("#44AA99", "#6699CC", "#882255","#999933"), guide = "none") +
  theme_dag()

basic_dag <- dagify(
  Y ~ X + A,
  X ~ A,
  B ~ X + Y,
  exposure = "X",
  outcome = "Y",
  coords = list(
    x = c(X = 1, Y = 3, A = 2, B = 2),
    y = c(X = 1, Y = 1, A = 2, B = 0)
  )
)

basic_dag_plot <- basic_dag |> 
  tidy_dagitty() |> 
  mutate(var_type = case_when(
    name == "X" ~ "Exposure",
    name == "Y" ~ "Outcome",
    str_detect(name, "A") ~ "Confounder",
    str_detect(name, "B") ~ "Collider"
  )) |> 
  mutate(var_label = case_match(name,
    "X" ~ "Treatment",
    "Y" ~ "Outcome",
    "A" ~ "Confounder",
    "B" ~ ""
  ))

ggplot(basic_dag_plot, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(aes(color = var_type), size = 12) +
  geom_richtext(
    aes(label = name), fontface = "bold", color = "white",
    fill = NA, label.color = NA,
    label.padding = grid::unit(c(3, 0, 0, 0), "pt")
  ) + 
  geom_dag_text(
    data = filter(basic_dag_plot, var_type != "Confounder"), aes(label = var_label), 
    nudge_y = -0.25, color = "black", size = 11, size.unit = "pt"
  ) +
  geom_dag_text(
    data = filter(basic_dag_plot, var_type == "Confounder"), aes(label = var_label), 
    nudge_y = 0.26, color = "black", size = 11, size.unit = "pt"
  ) +
   geom_text(
    aes(x = (x + xend) / 2, y = (y + yend + 0.4) / 2, label =  c("","","","","-1.72","")),
    color = "black", size = 5
  ) +
  ggtitle("Confounder \ny ~ x + a \n\n") +
  scale_color_manual(values = c("#44AA99", "#6699CC", "#882255","#999933"), guide = "none") +
  theme_dag()

basic_dag <- dagify(
  Y ~ X + A,
  X ~ A,
  B ~ X + Y,
  exposure = "X",
  outcome = "Y",
  coords = list(
    x = c(X = 1, Y = 3, A = 2, B = 2),
    y = c(X = 1, Y = 1, A = 2, B = 0)
  )
)

basic_dag_plot <- basic_dag |> 
  tidy_dagitty() |> 
  mutate(var_type = case_when(
    name == "X" ~ "Exposure",
    name == "Y" ~ "Outcome",
    str_detect(name, "A") ~ "Confounder",
    str_detect(name, "B") ~ "Collider"
  )) |> 
  mutate(var_label = case_match(name,
    "X" ~ "Treatment",
    "Y" ~ "Outcome",
    "A" ~ "",
    "B" ~ "Collider"
  ))

ggplot(basic_dag_plot, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(aes(color = var_type), size = 12) +
  geom_richtext(
    aes(label = name), fontface = "bold", color = "white",
    fill = NA, label.color = NA,
    label.padding = grid::unit(c(3, 0, 0, 0), "pt")
  ) + 
  geom_dag_text(
    data = filter(basic_dag_plot, var_type != "Confounder"), aes(label = var_label), 
    nudge_y = -0.25, color = "black", size = 11, size.unit = "pt"
  ) +
  geom_dag_text(
    data = filter(basic_dag_plot, var_type == "Confounder"), aes(label = var_label), 
    nudge_y = 0.26, color = "black", size = 11, size.unit = "pt"
  ) +
   geom_text(
    aes(x = (x + xend) / 2, y = (y + yend + 0.4) / 2, label =  c("","","","","0.78","")),
    color = "black", size = 5
  ) +
  ggtitle("Collider \ny ~ x + b \n\n") +
  scale_color_manual(values = c("#44AA99", "#6699CC", "#882255","#999933"), guide = "none") +
  theme_dag()
```

**Figure 2:** Various regression approaches to estimate the Y ~ X relationship (simulated value -1.71). Coefficients on arrows represent the estimated magnitude and direction of relationships. Conditioning on the confounder (A) returns the simulated value.


- Without consideration of the causal structure underlying data, it is easy to draw incorrect conclusions.
- Below is a simple example using a simulated dataset with four variables (Y, X, A, B).