---
title: "Generalised Linear Madness"
date: today
authors: "Tom Moore, James Bristow"
fig-pos: 'H'
toc: true
lof: true
lot: true
link-citations: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, output=TRUE, eval=TRUE}
library(dplyr)
library(stringr)
library(readxl)
library(here)
library(janitor)
library(marginaleffects)
library(pins)
library(reshape2)
library(arrow)
library(cowplot)
library(modelr)
library(bnlearn)
library(future)
library(dagitty)
library(cmdstanr)
library(brms)
library(ggdag)
library(tidybayes)
library(ggtext)
library(ggplot2)
library(simDAG)
library(rethinking)
library(forcats)

bayesplot::color_scheme_set("blue")                                   
options(mc.cores = 1)                           
```

# Introduction

This analysis demonstrates science before statistics by incorporating
prior knowledge in the modeling approach. Scientific phenomena are often
modeled as linear relationships (i.e., generalized linear (mixed)
models), which may not capture the data generating process.

Fitting non-linear models such as logarithmic, polynomial curves, or
splines may give us a good prediction of relationships, but doesn't
allow us to estimate parameters of interest (e.g., maximum value or
growth rate).

## Data and Methods

During my PhD, I measured the length and weight of \~3000 freshwater
mussels (yes I'm that exciting!). I want to describe the relationship
between length and weight between mussels collected inside macrophyte
beds (aquatic plant beds: treatment) vs not (control) based on how the
data was generated.

Although the length-weight relationship isn't technically a temporal
relationship (even I'm not going to measure \~3000 mussels repeatedly
over time), substituting time for space (length) will allow us to fit a
models that estimate the: a) maximum weight; and b) growth rate.

To examine the relationship between the length and weight of freshwater
mussels, generalized (non-)linear models were performed in R (R Core
Team 2022) using the brms package (Bürkner, 2017). Linear, Gompertz
(modified), and Logistic models were fitted with treatment (none vs
macrophyte) as a predictor variable. A Gamma distribution was selected
for the response since weight has positive and continuous values. Models
were compared using leave-one-out cross validation and model validation
was performed using prior and posterior predictive checks, as well as
effective samples sizes and r-hats (McElreath, 2020).

::: {.callout-tip title="Data" collapse="true"}
```{r glms-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

df <- readr::read_csv(here("pages/stan_workshop/data/Mussel_df_15.01.20.csv"), show_col_types = FALSE)

df <- df %>% dplyr::select(c(Length,Height,Width,Weight,Treatment,Macrophyte,Site))

df <- df[complete.cases(df),]  # Select only complete cases

head(df)
```
:::

::: panel-tabset
### Prior linear

```{r glms-m1-bad-priors, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

mod_formula <- bf(Weight ~ 0 + Intercept + Length + Treatment)
  
priors <- c(prior(class = b, coef = Intercept, normal(0,1)),       # Intercept
            prior(class = b, coef = TreatmentNone, normal(50,5)),  # Difference between treatments
            prior(class = b, normal(0,10), lb = -10, ub = -2)      # Slope 
            )

# Priors
nl_mod_1_priors <- brm(mod_formula, family = Gamma(link = "identity"),
             prior = priors, 
             sample_prior = "only", seed = 4,
             data = df,
             threads = threading(threads = NULL, grainsize = 1250, static = FALSE),
             backend = "cmdstanr", 
             file = here("pages", "stan_workshop", "model", "NL_mod_1_priors_bad"),
             silent = 0, refresh = 100,
             chains = 4, cores = 4, warmup = 1000, iter = 2000)

conditional_effects(nl_mod_1_priors, effect = "Length", spaghetti = T, ndraws = 100)
```

### Prior Gompertz

```{r glms-m2-good-priors, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
mod_formula <- bf(Weight ~ k * exp( -exp( -(r * (Length - delay)))),
                  k + r + delay ~ 0 + Intercept + Treatment,
                  nl = TRUE)
  
priors <- c(prior(nlpar = "k",     lb = 0, normal(50,20)),  # Max Weight 
            prior(nlpar = "r",     lb = 0, normal(0,0.05)), # Growth Rate
            prior(nlpar = "delay", lb = 0, normal(30,2))    # Delay across the x-axis 
            )

# Priors
nl_mod_2_priors <- brm(mod_formula, family = Gamma(link = "identity"),
             prior = priors, 
             sample_prior = "only", seed = 4,
             data = df, 
             threads = threading(threads = NULL, grainsize = 1250, static = FALSE),
             backend = "cmdstanr", 
             file = here("pages", "stan_workshop", "model", "NL_mod_2_priors_good"),
             silent = 1, refresh = 100,
             chains = 4, cores = 4, warmup = 1000, iter = 2000)

conditional_effects(nl_mod_2_priors, effect = "Length", spaghetti = T, ndraws = 100)
```

### Prior logistic

```{r glms-m3-good-priors, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

mod_formula <- bf(Weight ~ M / (1 + (M - mu0) / mu0 * exp( -b * Length)),
                  mu0 + b + M ~ 0 + Intercept + Treatment,
                  nl = TRUE)
  
priors <- c(prior(nlpar = "mu0", lb = 0, normal(0,0.1)),  # Weight at length zero
            prior(nlpar = "b",   lb = 0, normal(0,0.1)),  # Growth rate
            prior(nlpar = "M",   lb = 0, normal(50,20))   # Max Weight
            )

# Priors
nl_mod_3_priors <- brm(mod_formula, family = Gamma(link = "identity"),
             prior = priors, 
             sample_prior = "only", seed = 4,
             data = df,
             threads = threading(threads = NULL, grainsize = 1250, static = FALSE),
             backend = "cmdstanr", 
             file = here("pages", "stan_workshop", "model", "NL_mod_3_priors_good"),
             silent = 0, refresh = 100,
             chains = 4, cores = 4, warmup = 1000, iter = 2000)

conditional_effects(nl_mod_3_priors, effect = "Length", spaghetti = T, ndraws = 100)
```

### Models

```{r glms-logistic-priors, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| label: model_3
#| echo: true

fit_brm <- function(formula, model_name, priors) {
  brm(formula,
      family = Gamma(link = "identity"),
      data = df,
      prior = priors,
      backend = "cmdstanr",
      file = here("pages", "stan_workshop", "model", model_name),
      threads = threading(3),
      silent = 0,
      refresh = 500,
      chains = 4,
      cores = 4,
      warmup = 2000,
      iter = 4000)
}

# Linear
mod_formula <- bf(Weight ~ 0 + Intercept + Length + Treatment)     # Run with Gaussian family 
  
priors <- c(prior(class = b, coef = Intercept, normal(0,1)),       # Intercept
            prior(class = b, coef = TreatmentNone, normal(50,5)),  # Difference between treatments
            prior(class = b, normal(0,10), lb = -10, ub = -2)      # Slope 
            )

nl_mod_1 <- fit_brm(mod_formula, "NL_mod_1", priors)

# Gompertz
mod_formula <- bf(Weight ~ k * exp( -exp( -(r * (Length - delay)))),
                  k + r + delay ~ 0 + Intercept + Treatment,
                  nl = TRUE)
  
priors <- c(prior(nlpar = "k",     lb = 0, normal(50,20)),  # Max Weight 
            prior(nlpar = "r",     lb = 0, normal(0,0.05)), # Growth Rate
            prior(nlpar = "delay", lb = 0, normal(30,2))    # Delay across the x-axis 
            )

nl_mod_2 <- fit_brm(mod_formula, "NL_mod_2", priors)

# Logistic
mod_formula <- bf(Weight ~ M / (1 + (M - mu0) / mu0 * exp( -b * Length)),
                  mu0 + b + M ~ 0 + Intercept + Treatment,
                  nl = TRUE)
  
priors <- c(prior(nlpar = "mu0", lb = 0, normal(0,0.1)),  # Weight at length zero
            prior(nlpar = "b",   lb = 0, normal(0,0.1)),  # Growth rate
            prior(nlpar = "M",   lb = 0, normal(50,20))   # Max Weight
            )

nl_mod_3 <- fit_brm(mod_formula, "NL_mod_3", priors)
```

### Output

```{r glms-logistic-summary, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| label: model-output_3
#| echo: false
#| 
summary(nl_mod_1)
summary(nl_mod_2)
summary(nl_mod_3)
```

### Validation

```{r glms-logistic-ppc, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| label: validation_3
#| echo: false
#| message: false
#| fig-show: hold
#| results: false
#| layout-ncol: 2
#| fig.width: 4
#| fig.height: 2

pp_check(nl_mod_1)
pp_check(nl_mod_2)
pp_check(nl_mod_3)
```

### Plot linear

```{r glms-plot-linear, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

df %>%
  group_by(Treatment) %>%
  data_grid(Length = seq_range(Length, n = 101)) %>%
  add_epred_draws(nl_mod_1, ndraws = 100) %>%
  ggplot(aes(x = Length, y = Weight, color = Treatment)) +
  theme_classic() +
  geom_point(data = df, alpha = 0.25) +
  geom_line(aes(y = .epred, group = paste(Treatment, .draw)), alpha = .1, color = "grey50") +
  scale_x_continuous(limits = c(0,90), breaks = seq(0,90,10)) +
  scale_y_continuous(limits = c(-100,100), breaks = seq(-100,100,10)) +
  facet_wrap(~Treatment) +
  scale_color_brewer(palette = "Dark2")
```

### Plot Gompertz

```{r glms-plot-gompertz, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

df %>%
  group_by(Treatment) %>%
  data_grid(Length = seq_range(Length, n = 101)) %>%
  add_epred_draws(nl_mod_2, ndraws = 100) %>%
  ggplot(aes(x = Length, y = Weight, color = Treatment)) +
  theme_classic() +
  geom_point(data = df, alpha = 0.25) +
  geom_line(aes(y = .epred, group = paste(Treatment, .draw)), alpha = .1, color = "grey50") +
  scale_x_continuous(limits = c(0,90), breaks = seq(0,90,10)) +
  scale_y_continuous(limits = c(0,80), breaks = seq(0,80,10)) +
  facet_wrap(~Treatment) +
  scale_color_brewer(palette = "Dark2")
```

### Plot Logistic

```{r glms-plot-logistic, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

df %>%
  group_by(Treatment) %>%
  data_grid(Length = seq_range(Length, n = 101)) %>%
  add_epred_draws(nl_mod_3, ndraws = 100) %>%
  ggplot(aes(x = Length, y = Weight, color = Treatment)) +
  theme_classic() +
  geom_point(data = df, alpha = 0.25) +
  geom_line(aes(y = .epred, group = paste(Treatment, .draw)), alpha = .1, color = "grey50") +
  scale_x_continuous(limits = c(0,90), breaks = seq(0,90,10)) +
  scale_y_continuous(limits = c(0,80), breaks = seq(0,80,10)) +
  facet_wrap(~Treatment) +
  scale_color_brewer(palette = "Dark2")
```

### Model comparison

```{r glms-loo, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

loo(nl_mod_1, nl_mod_2, nl_mod_3)
```

### Stan code

```{r glms-stan-code, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| label: stan_3
#| echo: false

mod_formula <- bf(Weight ~ k * exp( -exp( -(r * (Length - delay)))),
                  k + r + delay ~ 0 + Intercept + Treatment,
                  nl = TRUE)
  
priors <- c(prior(nlpar = "k",     lb = 0, normal(50,20)),  # Max Weight 
            prior(nlpar = "r",     lb = 0, normal(0,0.05)), # Growth Rate
            prior(nlpar = "delay", lb = 0, normal(30,2))    # Delay across the x-axis 
            )

make_stancode(
  formula = mod_formula,
  data = df,
  prior = priors,
  family = Gamma(link = "identity")
)
```
:::

## Conclusion

Models can be used to estimate parameters of interest that better
capture data generating process based on ecological theory.

Prior information can be useful in constraining models to realistic
estimates.

Here the Gompertz and logistic models estimated the growth rate to be
0.03 and 0.10 and maximum weight to be 115 (g) and 50.0 (g),
respectively.

### Food for thought:

Which model is the best at capturing the data generating process?

Could ecological theory informed techniques be applied to your own
research?

# Beyond brms: in space

This section demonstrates the use of Stan to fit a choice model based on
the hidden minds and observed behavior data example from McElreath
(2020). The model estimates the probability of choosing a particular
option based on the majority and minority choices, as well as the
maverick and random choices.

The five stratergies are:

-   Follow the Majority: Copy the majority demonstrated color.
-   Follow the Minority: Copy the minority demonstrated color.
-   Maverick: Choose the color that no demonstrator chose.
-   Random: Choose a color at random, ignoring the demonstrators.
-   Follow First: Copy the color that was demonstrated first. This was
    either the majority color (when majority_first equals 1) or the
    minority color (when 0). (p. 533)

::: panel-tabset
```{r glms-stan-space-state, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
USE_VB <- FALSE

data(Boxes, package = "rethinking")

df_ss <- Boxes # hidden minds and observed behavior data example

# prep data 
dat_list <- list(
  N = nrow(df_ss),
  y = df_ss$y,
  majority_first = df_ss$majority_first)

choice_model_1 <- readLines(here("pages", "stan_workshop", "stan", "choice_model_1.stan"))

if(USE_VB) {
  choice_mod_1 <-
    rstan::vb(rstan::stan_model(model_code = choice_model_1), 
         data = dat_list
  )
} else {
  choice_mod_1 <- rstan::stan(model_code = choice_model_1,
       data = dat_list,
       iter = 2000,
       chains = 3,
       cores = 1
  )
}
```

### Stan - Model

```{r glms-stan-space-show-code, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| echo: false
#| results: asis

cat(readLines(here("pages", "stan_workshop", "stan", "choice_model_1.stan")), sep = "\n")
```

### Output

```{r glms-stan-space-state-results, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| label: model_output_3

print(choice_mod_1, pars = c("p"), digits = 3)
```

### Plot

```{r glms-stan-space-state-plot, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

#| label: plot_3
#| fig.height: 3

label <- c("Majority~(italic(s)[1])", 
           "Minority~(italic(s)[2])",
           "Maverick~(italic(s)[3])", 
           "Random~(italic(s)[4])", 
           "Follow~First~(italic(s)[5])")

rethinking::precis(choice_mod_1, depth = 2) %>% 
  data.frame() %>% 
  mutate(name = factor(label, levels = label)) %>% 
  mutate(name = fct_rev(name)) %>% 
  
  ggplot(aes(x = mean, xmin = X5.5., xmax = X94.5., y = name)) +
  theme_bw() +
  geom_vline(xintercept = .2, linewidth = 1/4, linetype = 2) +
  geom_pointrange(linewidth = 1/4, fatten = 6/4) +
  scale_x_continuous(expression(italic(p[s])), limits = c(0, 1),
                     breaks = 0:5 / 5) +
  scale_y_discrete(NULL, labels = ggplot2:::parse_safe) +
  theme(axis.text.y = element_text(hjust = 0))
```
:::

# References

McElreath, R. (2020). Statistical rethinking: A Bayesian course with
examples in R and Stan. Chapman and Hall/CRC.

R Core Team (2022). R: A language and environment for statistical
computing. R Foundation for Statistical Computing, Vienna, Austria.
<https://www.R-project.org/>

::: {.callout-tip title="Session Information" collapse="true"}
```{r session-info, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, output=TRUE, eval=TRUE}

#| label: session-info

utils::sessionInfo()
```
:::
