---
title: "MLOps using vetiver"
date: today
authors: "James Bristow, Tom Moore"
fig-pos: 'H'
toc: true
lof: true
lot: true
link-citations: true
editor: 
  markdown: 
    wrap: 72
execute:
  warning: false
  message: false
---

# Introduction

In this section, we will demonstrate how one may combine *vetiver* and
*plumber* with *Docker* and *Kubernetes* to enable Machine Learning
Operations (MLOps) practices. MLOps is the practice of combining machine
learning (ML) development with software engineering and operational
practices. It applies principles from DevOps like continuous
integration, continuous delivery (CI/CD), version control, and
monitoring to the lifecycle of ML models to make ML systems
reproducible, scalable, and reliable. Models can then be updated safely,
deployed efficiently, and monitored for performance drift over time

As a reminder:

- [vetiver: For model versioning, sharing, and deployment](https://rstudio.github.io/vetiver-r/)
- [plumber: For creating light-weight web applications, and deploying models as microservices](https://www.rplumber.io/)
- [pins: For uploading, downloading, and sharing resources.](https://pins.rstudio.com/)

We will elucidate upon *Docker* and *Kubernetes* in subsequent sections.

We first load all required dependencies.

```{r setup, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, output=TRUE, eval=TRUE}
Sys.setenv("S3_ENDPOINT_URL" = "http://127.0.0.1:9000")

box::use(../../R/proj_utils)

suppressWarnings(
  suppressPackageStartupMessages({
    library(rstan)
    library(dplyr)
    library(stringr)
    library(here)
    library(janitor)
    library(pins)
    library(vetiver)
    library(bayesplot)
    library(modelr)
    library(future)
    library(paws.common)
    library(cmdstanr)
    library(brms)
    library(paws.storage)
    library(tidybayes)
    library(plotly)
    library(ggplot2)
    library(simDAG)
    library(rethinking)
    library(forcats)
    library(htmlwidgets)
  })
)

source(here("R/stan_vetiver.R"))

bayesplot::color_scheme_set("blue")                                   
options(mc.cores = 1)                           
```

# Data

We will load mussel data collected during Tom Moore's PhD.

::: {.callout-tip title="Data" collapse="true"}
```{r stan-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

df <- readr::read_csv(here("pages/stan_workshop/data/Mussel_df_15.01.20.csv"), show_col_types = FALSE) |>
        dplyr::select(c(Length,Height,Width,Weight,Treatment,Macrophyte,Site))
df <- df[complete.cases(df),]  

head(df)
```
:::

# Model Fitting

We will fit the full Gompertz model.

```{r stan-gompertz, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}
#| label: gompertz
#| echo: true

fit_brm <- function(formula, model_name, priors) {
  brm(
    formula,
    family = Gamma(link = "identity"),
    data = df,
    prior = priors,
    backend = "cmdstanr",
    file = here("pages", "stan_workshop", "model", model_name),
    threads = threading(3),
    silent = 0,
    refresh = 500,
    chains = 4,
    cores = 4,
    warmup = 2000,
    iter = 4000
  )
}

mod_formula <- bf(Weight ~ k * exp(-exp(-(r * (
  Length - delay
)))), k + r + delay ~ 0 + Intercept + Treatment, nl = TRUE)

priors <- c(
  prior(nlpar = "k", lb = 0, normal(50, 20)),
  prior(nlpar = "r", lb = 0, normal(0, 0.05)),
  prior(nlpar = "delay", lb = 0, normal(30, 2)) 
)

gompertz_model <- fit_brm(mod_formula, "NL_mod_2", priors)
gompertz_model$training <- df |> select(-Weight)
gompertz_model$response <- df |> select(Weight)

gompertz_model
```

# Model versioning, sharing, and deployment

We will demonstrate how to perform model versioning, sharing, and
deployment using a combination of *pins*, *vetiver*, and *plumber*. [We'll store everything in S3. Amazon S3 (Simple Storage Service) is cloud storage for files, called "objects", organised in "buckets". It offers scalable, secure, and flexible storage with features like versioning, encryption, and programmatic access via URLs or AWS tools.](https://docs.aws.amazon.com/AmazonS3/latest/userguide/Welcome.html) In this example, we will first specify our S3 bucket.

```{r stan-s3-board, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
board <- board_s3(
  "data", 
  access_key = "user", 
  secret_access_key = "password", 
  region = "us-east-2", 
  endpoint = Sys.getenv("S3_ENDPOINT_URL") 
)
```

We will generate and save several plots for our Gompertz model to upload into an S3 bucket:

::: {.callout-tip title="Plots" collapse="true"}
```{r stan-vetiver-plots-save, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
save_s3_plot <- function(p, object_name, model, board) {
  outfile <- paste0(object_name, ".html")

  p |> 
    ggplotly() |>
    saveWidget(file.path("/tmp", outfile))
    
  model <- proj_utils$model_s3_upload(
    model, board, Sys.getenv("S3_ENDPOINT_URL"), 
    "/tmp", outfile, object_name
  )
  model
}

gompertz_model <- pp_check(gompertz_model) |>
  save_s3_plot("gompertz_ppc", gompertz_model, board)

gompertz_model <- as_draws_df(gompertz_model) |>
  mcmc_dens_overlay() |>
  save_s3_plot("gompertz_mcmc_dens", gompertz_model, board)

gompertz_model <- as_draws_df(gompertz_model) |>
  mcmc_trace() |>
  save_s3_plot("gompertz_mcmc_trace", gompertz_model, board)

p <- ppc_intervals(
  y = df$Weight,
  yrep = posterior_predict(gompertz_model, ndraws = 50),
  x = df$Length
) +
  labs(x = "Length", y = "Weight") 
gompertz_model <- save_s3_plot(p, "gompertz_ppc_intervals", gompertz_model, board)
```
:::

We will next save our Gompertz model to an S3 bucket using *pins* and *vetiver*.

```{r stan-vetiver-save, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
v_model <- vetiver_model(gompertz_model, "gompertz")
vetiver_pin_write(v_model, board = board)
```

We will next load and interact with the saved Gompertz model locally using a plumber web application.

```{r stan-vetiver-load, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=FALSE}
loaded_v <- board |> 
  vetiver_pin_read("gompertz")

pr(here("app", "api", "plumber.R")) |>
  vetiver_api(loaded_v) |>
  pr_run(port = 8088)
```

# Docker

[Docker is a containerisation technology that allows developers to package an application along with all its dependencies, libraries, and configuration files into a single unit called a container](https://www.docker.com/). This ensures that the application runs consistently across different environments, whether on a developer's laptop, a testing server, or a production cloud environment. Docker helps developers "build once, run anywhere" without worrying about environment mismatches.

Our objective is to package up our plumber web application and model within a Docker container for subsequent deployment. In brief:

1. [We have a Dockerfile which contains a sequence of steps needed to build a Docker image. We copy our code from the R workshop repo into the Docker image (including the plumber web application code).](https://github.com/Plant-Food-Research-Open/r-pipeline-development-workshop/blob/main/dockerfiles/shiny/Dockerfile.0.1.0)
2. [We define a pipeline that builds the Docker image and pushes it into a container repository.](https://github.com/Plant-Food-Research-Open/r-pipeline-development-workshop/blob/main/.github/workflows/build-r.yaml)
3. [Our stored image may be found in the GitHub container repository.](https://github.com/Plant-Food-Research-Open/r-pipeline-development-workshop/pkgs/container/shiny-rpipelines)
4. We can then pull down our stored image and instantiate a container to deploy our plumber web application, either using Docker directly or with a container orchestration platform such as Kubernetes.
5. Our model can then be loaded into the web application within the container via S3.

# Kubernetes

[Kubernetes is an open-source platform that automates the deployment, scaling, and management of containerised applications, such as our deployed models.](https://kubernetes.io/)

It organizes containers into logical units called pods, schedules them across a cluster of machines, and handles networking, storage, and load balancing. This ensures that your applications run reliably even as demand changes or failures occur. Kubernetes also provides self-healing features, automatically restarting or rescheduling containers when they fail. 

In brief:

1. Kubernetes will pull down our Docker image containing our plumber web application. 
2. Kubernetes will then deploy our web application within a pod inside our Kubernetes cluster.
3. Our S3 server is also running on Kubernetes, allowing us to load our model and any other artifacts of interest.
