---
title: "MLOps using vetiver"
date: today
authors: "James Bristow, Tom Moore"
fig-pos: 'H'
toc: true
lof: true
lot: true
link-citations: true
editor: 
  markdown: 
    wrap: 72
execute:
  warning: false
  message: false
---

# Introduction

In this section, we will demonstrate how one may combine *vetiver* and
*plumber* with *Docker* and *Kubernetes* to enable Machine Learning
Operations (MLOps) practices. MLOps is the practice of combining machine
learning (ML) development with software engineering and operational
practices. It applies principles from DevOps—like continuous
integration, continuous delivery (CI/CD), version control, and
monitoring to the lifecycle of ML models to make ML systems
reproducible, scalable, and reliable. Models can then be updated safely,
deployed efficiently, and monitored for performance drift over time

As a reminder:

-   [vetiver: For model versioning, sharing, and
    deployment](https://rstudio.github.io/vetiver-r/)
-   [plumber: For creating light-weight web applications, and deploying
    models as microservices](https://www.rplumber.io/)

We will elucidate upon *Docker* and *Kubernetes* in subsequent sections.

We first load all required dependencies.

```{r setup, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE, output=TRUE, eval=TRUE}
suppressWarnings(
  suppressPackageStartupMessages({
    library(dplyr)
    library(stringr)
    library(here)
    library(janitor)
    library(pins)
    library(vetiver)
    library(modelr)
    library(future)
    library(paws.common)
    library(cmdstanr)
    library(brms)
    library(paws.storage)
    library(tidybayes)
    library(plotly)
    library(ggplot2)
    library(simDAG)
    library(rethinking)
    library(forcats)
  })
)

bayesplot::color_scheme_set("blue")                                   
options(mc.cores = 1)                           
```

## Data and Methods

We will load mussel data collected during Tom Moore's PhD.

::: {.callout-tip title="Data" collapse="true"}
```{r stan-data, include = TRUE, echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, output=TRUE, eval=TRUE}

df <- readr::read_csv(here("pages/stan_workshop/data/Mussel_df_15.01.20.csv"), show_col_types = FALSE) |>
        dplyr::select(c(Length,Height,Width,Weight,Treatment,Macrophyte,Site))
df <- df[complete.cases(df),]  

head(df)
```
:::
